---
title: "Introduction to transferegovr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to transferegovr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

TransfereGov publishes four open data APIs on
`api-publica.transferegov.gestao.gov.br`, covering seventy-four tables and about
7.5 million rows between them. This vignette shows how to find your way around
them and retrieve data without downloading more than you meant to.

The code here is not run when the vignette is built, because it would call the
government's servers.

```{r}
library(transferegovr)
```

## The four modules

```{r}
tg_modules()
#> # A tibble: 4 × 5
#>   module      label                  tables max_page_size url
#>   <chr>       <chr>                   <int>         <int> <chr>
#> 1 especiais   Special transfers          23           200 https://api-publ…
#> 2 fundoafundo Fund-to-fund transfers     20          1000 https://api-publ…
#> 3 parcerias   Partnerships               17           200 https://api-publ…
#> 4 ted         Decentralized credit       14          1000 https://api-publ…
```

`especiais` covers **special transfers**, the mechanism created by
Constitutional Amendment 105/2019 that lets an individual parliamentary
amendment send money straight to a municipality's account, with no agreement to
sign. `fundoafundo` covers **fund-to-fund transfers**, from a federal fund to a
state or municipal one — health, social assistance, education. `parcerias`
covers **partnerships** with civil society organizations, from the program
that announces them through to the bank statement of the account they are paid
from. `ted` covers **decentralized credit**, the *termos de execução
descentralizada* through which one federal body passes budget to another to
carry out a program on its behalf.

What is *not* here: SICONV agreement data, which is published as CSV downloads
rather than as an API.

## Finding a table

```{r}
tg_tables("parcerias")
#> # A tibble: 17 × 6
#>    module    table                 path                  columns params
#>    <chr>     <chr>                 <chr>                   <int>  <int>
#>  1 parcerias analise_proposta      analise-proposta            7      5
#>  2 parcerias beneficiario_emenda_… beneficiario_emenda_…      16     14
#>  3 parcerias cronograma_desembolso cronograma-desembolso       7      7
#>  …
```

`table` is the name to pass to `tg_get()`; `path` is the endpoint it maps to.
The two differ wherever the endpoint uses a hyphen, and either spelling is
accepted.

`tg_fields()` describes the columns, and `tg_params()` the filters:

```{r}
tg_fields("parcerias", "proposta")
#> # A tibble: 43 × 5
#>    field               r_type    api_type nested description
#>    <chr>               <chr>     <chr>    <chr>  <chr>
#>  1 id_proposta         double    integer  NA     Identificador único da prop…
#>  2 id_programa         double    integer  NA     Identificador do programa a…
#>  …

tg_params("parcerias", "proposta")
```

These two are not the same set. A column is what comes back; a parameter is what
you can filter on. Most columns are both, but not all.

Both work offline: the schema is frozen into the package from the APIs' own
OpenAPI documents. `tg_schema_date()` reports when.

## Retrieving rows

Each filter is named after a parameter, and parameters combine with AND:

```{r}
propostas <- tg_get(
  "parcerias", "proposta",
  sg_uf_recebedor = "PE",
  situacao_proposta = "Aprovada",
  .limit = 100
)
```

That is almost the entire filtering vocabulary. These services compare for
equality — there is no greater-than and no pattern match — and they publish no
ordering or column-selection parameter.

The exception is on identifiers. Some parameters take several values in one
request and match any of them. `tg_params()` marks them as `multiple`, with the
most each accepts — 100 in `especiais`, 200 elsewhere:

```{r}
params <- tg_params("parcerias", "parceria")
params[params$multiple, c("param", "max_values")]
#> # A tibble: 2 × 2
#>   param       max_values
#>   <chr>            <int>
#> 1 id_parceria        200
#> 2 id_proposta        200

tg_get("parcerias", "parceria", id_proposta = c(1, 2))
```

The OpenAPI documents do not say which parameters these are, and the names do
not either, so the package asked the service when it froze the schema.

For any other parameter, query each value and bind:

```{r}
library(purrr)

nordeste <- c("PE", "PB", "AL", "RN", "CE", "SE", "BA", "PI", "MA")

propostas <- list_rbind(map(
  nordeste,
  \(uf) tg_get("parcerias", "proposta", sg_uf_recebedor = uf, .limit = Inf)
))
```

### Enumerated parameters

Many parameters accept only a fixed set of values, and the package knows which:

```{r}
params <- tg_params("parcerias", "proposta")
params[lengths(params$values) > 0, c("param", "values")]
#> # A tibble: 5 × 2
#>   param                  values
#>   <chr>                  <list>
#> 1 sg_uf_recebedor        <chr [27]>
#> 2 situacao_proposta      <chr [5]>
#> 3 in_situacao_analise    <chr [4]>
#> …

params$values[[match("situacao_proposta", params$param)]]
#> [1] "Em Análise"    "Rejeitada"     "Aprovada"      "Em Elaboração"
#> [5] "Inativada"
```

A value outside the set fails before the request is made:

```{r}
tg_count("parcerias", "proposta", situacao_proposta = "Aprovado")
#> Error in `tg_count()`:
#> ! "Aprovado" is not a permitted value for `situacao_proposta`.
#> ℹ Did you mean "Aprovada"?
#> ℹ It accepts "Em Análise", "Rejeitada", "Aprovada", "Em Elaboração", and
#>   "Inativada".
```

## Why parameter names are checked

This is the one thing worth internalizing about these APIs.

**They ignore a query parameter they do not recognize.** No warning, no 400 —
the request succeeds and returns the unfiltered table. The parameter on
`/proposta` is `situacao_proposta`; write `in_situacao_proposta`, which is what
the sibling `/parceria` endpoint calls its own version, and you get every one of
the 89,415 proposals instead of the 85,041 that are approved. Nothing in the
response says so.

So the package refuses to send a name the frozen schema does not know:

```{r}
tg_count("parcerias", "proposta", in_situacao_proposta = "Aprovada")
#> Error in `tg_count()`:
#> ! Unknown filter: "in_situacao_proposta".
#> ✖ The API ignores a parameter it does not recognize and returns every row, so
#>   this would look like a query that matched nothing in particular.
#> ℹ Did you mean "situacao_proposta"?
```

If the API gains a parameter after the packaged schema was built, turn the check
off with `options(transferegovr.validate = FALSE)` — and know what you are
trading away.

For the same reason a repeated parameter is refused rather than sent: these
services keep the last occurrence and discard the rest silently, so
`sg_uf_recebedor = "PE", sg_uf_recebedor = "PB"` would quietly mean `"PB"`.

## Types

Columns are typed from the frozen schema rather than inferred from the values,
so a column that happens to be entirely null on one page does not come back
logical while the next page returns it as character.

```{r}
propostas <- tg_get("parcerias", "proposta", .limit = 5)

class(propostas$dt_proposta)
#> [1] "Date"
class(propostas$vl_total_planejamento_gastos)
#> [1] "numeric"
```

Integers are returned as double. These documents declare no `format`, so int32
and int64 cannot be told apart, and identifiers here genuinely exceed
`.Machine$integer.max` — `cd_parceria` reaches 202500037062, which as an integer
would be `NA`.

### List columns

Some tables have no endpoint of their own: the API folds them into their parent
as an array. Those arrive as list columns.

```{r}
programas <- tg_get("parcerias", "programa", .limit = 20)

fields <- tg_fields("parcerias", "programa")
fields$field[!is.na(fields$nested)]
#> [1] "ufs_habilitadas"      "programa_atende_a"    "categorias_despesa"
#> [4] "resultados_esperados" "indicadores_programa"

tg_fields("parcerias", "programa", nested = "ufs_habilitadas")
#> # A tibble: 3 × 5
#>   field   r_type    api_type nested description
#>   <chr>   <chr>     <chr>    <chr>  <chr>
#> 1 nm_uf   character string   NA     NA
#> 2 sg_uf   character string   NA     NA
#> 3 cd_ibge double    integer  NA     NA
```

To flatten one:

```{r}
library(dplyr)
library(tidyr)

programas |>
  select(id_programa, ufs_habilitadas) |>
  unnest_longer(ufs_habilitadas) |>
  unnest_wider(ufs_habilitadas)
```

There are 5 such columns in `fundoafundo` and 13 in `parcerias`; `especiais` has
none.

## How fresh is the data

Each module reports when it was last loaded. It is the only freshness signal
these APIs give — they send no `ETag`, `Cache-Control` or `Last-Modified`
header, which is also why the package caches responses itself rather than
relying on HTTP caching.

```{r}
tg_updated_at("parcerias")
#> [1] "2026-08-03 UTC"
```

## Where to go next

* `vignette("pagination")` — collecting a large table without losing rows.
* `vignette("joining-tables")` — how the tables fit together.
